Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2016-10-19 paper.tex
Add todo: code up the moment integrals
Remove grouped address setting for Revtex 4.1
Reduce scope of paper to simply including elastic scattering

\documentclass[aps,prd,twocolumn,superscriptaddress,groupedaddress]{revtex4-1}
% Intended to be built with pdflatex, using the REVTEX 4 package.
% Template taken from FermiLab APS Jouranl template.

@@

\title{Scattering and Pair Processes in General Relativistic Ray Tracing for Neutrinos}
\todo{you sure about including all processes?}

@@

\begin{abstract}
  We present a covariant ray tracing algorithm for computing high-resolution
  neutrino distributions in general numerical spacetimes with hydrodynamical
  sources.
  Our formulation treats scattering and pair processes
  by incorporating estimates of the background neutrino fields.
  These background fields may be taken from a low-order moment simulation,
  be supplied analytically in simple cases,
  or be ignored entirely, in which case the method
  reduces to a standard state-of-the-art ray tracing formulation.
  It handles radiation in regimes spanning optically thick to optically thin.
  We test the new code, highlight its strengths and weaknesses, and
  apply it to simulations of neutron star mergers to
  1) compute neutrino fluxes as a function of energy and emission angle, and
  2) compute the radiation pressure tensor in the funnel of the disk,
  demonstrating a way to improve the closure relations used in the truncated
  moment formalism.
\end{abstract}

@@

\section{Introduction}
Neutrinos are one of the dominant energy transport phenomena at play in
neutron star mergers, heating, cooling, and pushing the disrupted nuclear
matter.
In addition, they change the composition of the matter via charged current
interactions.
Because neutrinos scatter over length scales both large and small with
respect to fluid scales, accurate models require a neutrino treatment that
respects the freedom of neutrino distribution functions to vary
drastically from thermodynamic equilibrium.

@@

We build upon existing ray tracing formulations
by eschewing any assumptions about the spacetime geometry
and by including scattering and pair processes in the integration along each
geodesic.
To incorporate these neutrino processes in our method, we employ
estimates of the first two angular moments of the distribution function
computed in a truncated moment scheme (formulated by
\cite{thor1981-truncated_moment, shib2011-truncated_moment}
and applied e.g. in \cite{ramp2002-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}).
If the first two angular moments are not available either from
a truncated moment evolution or a trustworthy analytical estimate,
our method reduces to the current state-of-the-art ray tracing method,
neglecting scattering and pair processes.

@@

\subsubsection*{Why do scattering and pair processes matter?}
Inelastic scattering and pair processes introduce significant changes to
a spectrum,
\todo{clarify these terms}
\todo{justify these statements}
and elastic scattering can signicantly modify neutrino
distributions in angle, and dilute the emitted spectrum over a larger emitting
surface \citep{pere2016-asl}, especially for heavy-lepton neutrinos.
Any phenomena that involve neutrino-neutrino interactions (neutrino oscillations,
and neutrino-antineutrino annihilation) can depend sensitively on the angular
distribution.
And spectral changes can strongly affect the nucleosynthesis
occuring in the ejected and irradiated material.
\todo{justify these statements}

@@

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
\todo{clarify how momenta are decoupled}
In this paper we outwit this limitation by incorporating coupling
terms that depend on either previously-evolved or analytical
estimates of the neutrino fields.
We also retain the freedom to drop the coupling terms, in which case our method
becomes similar to existing state-of-the-art ray tracing methods.
Our method is not a standalone radiation transport scheme,
but serves as the final component of a hybrid scheme,
piggy-backing on a lower-order radiation transport method as a
post-processing step.

@@

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \varepsilon^3 \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \varepsilon^3 \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \varepsilon^3 \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell^\mu \ell^\gamma,
\end{align}
representing specific energy density, specific energy flux, and
specific momentum flux, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
\todo{relate moments to other frames?}
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos\upsilon_1)\,d\upsilon_2$.
\todo{consistent with Eqn.~\ref{eqn:def_direction}?}

@@

We follow
\cite{brue1985-core_collapse} and \cite[Sec.~4]{shib2011-truncated_moment}
by separating these processes into four categories:
\begin{equation}
  \label{eqn:four_sources}
  C[f] \equiv C_{\rm AE} + C_{\rm SE} + C_{\rm SI} + C_{\rm TP},
\end{equation}
representing charged-current absorption and emission, elastic scattering,
inelastic scattering, and thermal pair annihilation and production. Each term
is computed by summing the weak interaction rates of the processes from the
relevant category given in Tab.~\ref{tab:neutrino_processes}.
These rates are computed most easily as traditional emissivities and opacities
in the fluid frame of reference;
but since we use them in the form of an invariant emissivity and opacity
(defined e.g. in \ref{eqn:inv_emissivity} and \ref{eqn:inv_opacity},
and obeying Eqn.~\ref{eqn:boltzmann_linear})
they take the same numerical value in any frame of reference
and are completely independent of our choice of fiducial observer $u^\alpha$.
\todo{confirm}

>>>>

\documentclass[aps,prd,twocolumn,superscriptaddress]{revtex4-1}
% Intended to be built with pdflatex, using the REVTEX 4.1 package.
% Template taken from FermiLab APS Jouranl template.

@@

\title{Elastic Scattering in General Relativistic Ray Tracing for Neutrinos}

@@

\begin{abstract}
  We present a covariant ray tracing algorithm for computing high-resolution
  neutrino distributions in general numerical spacetimes with hydrodynamical
  sources.
  Our formulation treats elastic scattering of neutrinos (changing direction
  but not energy) off of nuclei and
  nucleons by incorporating estimates of the background neutrino fields.
  These background fields may be taken from a low-order moment simulation,
  be supplied analytically in simple cases,
  or be ignored entirely, in which case the method
  reduces to a standard state-of-the-art ray tracing formulation.
  It handles radiation in regimes spanning optically thick to optically thin.
  We test the new code, highlight its strengths and weaknesses, and
  apply it to simulations of neutron star mergers to
  1) compute neutrino fluxes as a function of energy and emission angle, and
  2) compute the radiation pressure tensor in the funnel of the disk,
  demonstrating a way to improve the closure relations used in the truncated
  moment formalism.
\end{abstract}

@@

\section{Introduction}
Neutrinos are one of the dominant energy transport phenomena at play in
neutron star mergers, heating, cooling, and pushing the disrupted nuclear
matter.
In addition, they change the composition of the matter via charged current
interactions.
Because neutrinos scatter over length scales both large and small with
respect to fluid scales, accurate models require a neutrino treatment that
respects the freedom of neutrino distribution functions to vary
drastically from simple geometric distributions in thermodynamic equilibrium.

@@

We build upon existing ray tracing formulations
by eschewing any assumptions about the spacetime geometry
and by including elastic scattering in the integration along each
geodesic.
To incorporate scattering processes in our method, we employ estimates of the
lowest-order moment of the neutrino distribution function
computed in a truncated moment scheme (formulated by
\cite{thor1981-truncated_moment, shib2011-truncated_moment}
and applied e.g. in \cite{ramp2002-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}).
If moments are not available either from
a truncated moment evolution or a trustworthy analytical estimate,
our method reduces to the current state-of-the-art ray tracing method,
neglecting scattering.

@@

\subsubsection*{Why do scattering and pair processes matter?}
Elastic scattering (see Tab.~\ref{tab:neutrino_processes}) can signicantly
modify neutrino distributions in angle and dilute the emitted spectrum over a
larger emitting surface \citep{pere2016-asl}.
This is especially pronounced in the case of heavy-lepton neutrinos.
\todo{cite}
Inelastic scattering and pair processes can introduce further modifications.
Any phenomena that involve neutrino-neutrino interactions (neutrino oscillations,
and neutrino-antineutrino annihilation) can depend sensitively on the angular
distribution.
And spectral changes can strongly affect the nucleosynthesis
occuring in the ejected and irradiated material.
\todo{justify these statements}
The ray tracing method we present in this paper captures the dominant effects
of elastic scattering, while leaving the physics of inelastic scattering and
pair processes for later work.

@@

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
\todo{clarify how momenta are decoupled}
When we're solving for $f^{\nu_\sigma}(x^\alpha;p_\beta)$, we have no
information about $f^{\nu_\sigma}(x^\alpha;p'_\beta)$, and
$f^{\bar{\nu}_\sigma}(x^\alpha;p'_\beta)$, which are essential ingredients
for the source terms of the Boltzmann Equation describing the creation and
destruction of neutrinos due to the interaction processes described in
Tab.~\ref{tab:neutrino_processes}.

In this paper we outwit this limitation by incorporating coupled source terms
that depend on either previously-evolved or analytical
estimates of the neutrino fields.
We also retain the freedom to drop the coupling terms, in which case our method
becomes similar to existing state-of-the-art ray tracing methods.
Our method is not a standalone radiation transport scheme,
but serves as the final component of a hybrid scheme,
piggy-backing on a lower-order radiation transport method as a
post-processing step.

@@

\subsection{Moments of the Distribution Function}
\label{ssec:moments}
\todo{code these up}
We may take angular moments of the distribution function:
\begin{align}
  \label{eqn:J}
  J(\varepsilon) &=
  \varepsilon^3 \oint d\Omega' f(\varepsilon, \ell'_\beta) \\
  \label{eqn:Ha}
  H^\mu(\varepsilon) &=
  \varepsilon^3 \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell^\mu \\
  \label{eqn:Sab}
  S^{\mu\gamma}(\varepsilon) &=
  \varepsilon^3 \oint d\Omega' f(\varepsilon, \ell'_\beta) \ell^\mu \ell^\gamma,
\end{align}
representing specific energy density, specific energy flux, and
specific momentum flux, respectively.
Here ``specific'' refers to the quantity being integrable over neutrino energy.
\todo{relate moments to other frames?}
Integrals~\ref{eqn:J}--\ref{eqn:Sab} are performed over a solid angle in
momentum space while holding $\varepsilon$ constant:
$d\Omega \equiv d(\cos\upsilon_1)\,d\upsilon_2$.
\todo{consistent with Eqn.~\ref{eqn:def_direction}?}

@@

We follow
\cite{brue1985-core_collapse} and \cite[Sec.~4]{shib2011-truncated_moment}
by separating these processes into four categories:
\begin{equation}
  \label{eqn:four_sources}
  C[f] \equiv C_{\rm AE} + C_{\rm SE} + C_{\rm SI} + C_{\rm TP},
\end{equation}
representing charged-current absorption and emission, elastic scattering,
inelastic scattering, and thermal pair annihilation and production.
We present formulae for all of these processes, though in this work, we only
treat absorption/emission and elastic scattering.
Each term
is computed by summing the weak interaction rates of the processes from the
relevant category given in Tab.~\ref{tab:neutrino_processes}.
These rates are computed most easily as traditional emissivities and opacities
in the fluid frame of reference;
but since we use them in the form of an invariant emissivity and opacity
(defined e.g. in \ref{eqn:inv_emissivity} and \ref{eqn:inv_opacity},
and obeying Eqn.~\ref{eqn:boltzmann_linear})
they take the same numerical value in any frame of reference
and are completely independent of our choice of fiducial observer $u^\alpha$.
\todo{confirm}
